---
title: "lilyanalysis"
output: html_document
---

```{r load, include = FALSE}
library(readr)
library(dplyr)
library(tidyr)
library(tidyverse)
library(readxl)
library('corrr')
library(ggcorrplot)
library(factoextra)
library("FactoMineR")
library(missMDA)
library(wesanderson)
library("ggfortify")
library(ggcorrplot)
```

```{r read}
file_path <- "~/sapphire/data/serum_vit_D_study_with_lab_results.xlsx" 
data_summer <- read_excel(file_path, sheet = "ScreeningDataCollectionSummer")
data_winter <- read_excel(file_path, sheet = "ScreeningDataCollectionWinter")
data_6weeks <- read_excel(file_path, sheet = "ScreeningDataCollection6Weeks")
sun_expos <- read.csv("~/sapphire/data/sun_expos_data/sun_expos_long.csv")
sun_expos_summer <- sun_expos[sun_expos$collection_period == 'Summer', ]
sun_expos_winter <- sun_expos[sun_expos$collection_period == 'Winter', ]
sun_expos_6Weeks <- sun_expos[sun_expos$collection_period == '6Weeks', ]
```

# part 0: seasonal data cleaning

```{r join_seasonal}
summer_data <- left_join(data_summer, sun_expos_summer, 
                     by = c("ParticipantCentreID" = "participant_centre_id"))

winter_data <- left_join(data_winter, sun_expos_winter, 
                     by = c("ParticipantCentreID" = "participant_centre_id"))

six_week_data <- left_join(data_6weeks, sun_expos_6Weeks, 
                     by = c("ParticipantCentreID" = "participant_centre_id"))
```

```{r cleaning_seasonal}
# just subsetting some things I know I don't need 
#this ethnicity col was all NA so I'm making my own
summer_data = subset(summer_data, select = -c(Supplements, Medications, 
                                              EthnicitySpecifyOther, SmokingComments,
                                              x9if_apply_sunscreen_spf_used))
winter_data = subset(winter_data, select = -c(Supplements, Medications, 
                                              EthnicitySpecifyOther, SmokingComments, 
                                              ContinuedInStudy, IfNotContinuedInStudyReason,
                                              x9if_apply_sunscreen_spf_used))
six_week_data = subset(six_week_data, select = -c(Supplements, Medications, 
                                              EthnicitySpecifyOther, SmokingComments, 
                                              ContinuedInStudy, IfNotContinuedInStudyReason,
                                              x9if_apply_sunscreen_spf_used))
six_week_data = six_week_data[,!grepl("IfNoReasonForExclusion:",names(six_week_data))]
winter_data = winter_data[,!grepl("IfNoReasonForExclusion:",names(winter_data))]
summer_data = summer_data[,!grepl("IfNoReasonForExclusion:",names(summer_data))]
six_week_data = six_week_data[,!grepl("Req Num",names(six_week_data))]
winter_data = winter_data[,!grepl("Req Num",names(winter_data))]
summer_data = summer_data[,!grepl("Req Num",names(summer_data))]
```

```{r ethnicity_init}

ethnicity <- function(EthnicityAfricanBlack, EthnicityColoured, EthnicityWhite, 
                      EthnicityIndianAsian) {
  case_when(
    EthnicityAfricanBlack == TRUE & 
      EthnicityColoured == FALSE & 
      EthnicityWhite == FALSE & 
      EthnicityIndianAsian == FALSE ~ "Xhosa",
    EthnicityAfricanBlack == FALSE & 
      EthnicityColoured == TRUE & 
      EthnicityWhite == FALSE & 
      EthnicityIndianAsian == FALSE ~ "Cape_colored",
    TRUE ~ NA_character_
  )
}

summer_data <- summer_data %>%
  mutate(Ethnicity = ethnicity(EthnicityAfricanBlack, EthnicityColoured, 
                               EthnicityWhite, EthnicityIndianAsian))
summer_data = subset(summer_data, select = -c(EthnicityAfricanBlack, 
                                               EthnicityColoured, EthnicityWhite, 
                                               EthnicityIndianAsian))

winter_data <- winter_data %>%
  mutate(Ethnicity = ethnicity(EthnicityAfricanBlack, EthnicityColoured, 
                               EthnicityWhite, EthnicityIndianAsian))
winter_data = subset(winter_data, select = -c(EthnicityAfricanBlack, 
                                               EthnicityColoured, EthnicityWhite, 
                                               EthnicityIndianAsian))

six_week_data <- six_week_data %>%
  mutate(Ethnicity = ethnicity(EthnicityAfricanBlack, EthnicityColoured, 
                               EthnicityWhite, EthnicityIndianAsian))
six_week_data = subset(six_week_data, select = -c(EthnicityAfricanBlack, 
                                               EthnicityColoured, EthnicityWhite, 
                                               EthnicityIndianAsian))
```

```{r reflectance_metrics}
# regex that I hopefully don't have to use?
# site = sub("SkinReflectance(.*?)[EMRGBL\\*ab\\*]+\\d", "\\1", Variable)
# metric = sub(".*?([EMRGBL\\*ab\\*]+)\\d", "\\1", Variable)

sites <- c("Forehead", "RightUpperInnerArm", "LeftUpperInnerArm")
metrics <- c("E", "M", "R", "G", "B", "L\\*", "a\\*", "b\\*")

for (site in sites) {
  for (metric in metrics) {
    six_week_data <- six_week_data %>%
      mutate(
        !!paste0("SkinReflectance", site, metric, "_Avg") := rowMeans(
          select(., matches(paste0("SkinReflectance", site, metric, "[123]"))),
          na.rm = TRUE
        )
      )
  }
}

for (site in sites) {
  for (metric in metrics) {
    winter_data <- winter_data %>%
      mutate(
        !!paste0("SkinReflectance", site, metric, "_Avg") := rowMeans(
          select(., matches(paste0("SkinReflectance", site, metric, "[123]"))),
          na.rm = TRUE
        )
      )
  }
}

for (site in sites) {
  for (metric in metrics) {
    summer_data <- summer_data %>%
      mutate(
        !!paste0("SkinReflectance", site, metric, "_Avg") := rowMeans(
          select(., matches(paste0("SkinReflectance", site, metric, "[123]"))),
          na.rm = TRUE
        )
      )
  }
}

# now remove columns that end in metric[123] (the not averaged measurements)
winter_data <- winter_data %>%
  select(-matches(".*[EMRGBL\\*a\\*b\\*]\\d$"))

summer_data <- summer_data %>%
  select(-matches(".*[EMRGBL\\*a\\*b\\*]\\d$"))

six_week_data <- six_week_data %>%
  select(-matches(".*[EMRGBL\\*a\\*b\\*]\\d$"))
```

# part 1: joining

First step is to pivot data to get all numeric data together.

```{r joining}
six_week_rename <- six_week_data %>%
    rename_with(~ paste0(., "-6weeks"), -ParticipantCentreID)

joinone <- left_join(summer_data, winter_data, 
                     by = "ParticipantCentreID", 
                     suffix = c("-summer", "-winter"))

joined_data <- left_join(joinone, six_week_rename, 
                         by = "ParticipantCentreID")

head(joined_data)
```

```{r pivot_longer}
long_data <- joined_data %>%
  pivot_longer(
    cols = -ParticipantCentreID,  
    names_to = c(".value", "Season"),        # Split column names into two parts
    names_sep = "-"                          # Separate at _
  )
```

Now we have a wide and long dataset that combines the data across summer, winter, and 6 week.

# part 0.5: initial visualization

```{r initial_visuals}
ggplot(long_data, aes(VitDResult, SkinReflectanceForeheadM_Avg, col = Ethnicity, fill = Ethnicity)) + 
  geom_point(aes(shape = Season), size = 3) 
```

# part 1: seasonal PCA

```{r summer}
omittedsummer <- na.omit(summer_data)
summer <- select_if(omittedsummer, is.numeric)
summer <- subset(summer, select = -Gender)
summer <- scale(summer)
summerpca <- prcomp(summer)
summary(summerpca)
summerpca$loadings[, 1:2]

fviz_eig(summerpca, addlabels = TRUE)

fviz_pca_var(summerpca, col.var = "black")

fviz_cos2(summerpca, choice = "var", axes = 1:2)
```

From all of the contributions we know that dimension 1, which accounts for 67.5% of the variance in the summer data, is made up of all of the measures of pigmentation. This is pretty consistent across all of the PCAs. 

```{r winter}
omittedwinter <- na.omit(winter_data)
winter <- select_if(omittedwinter, is.numeric)
winter <- subset(winter, select = -Gender)
winter <- scale(winter)
winterpca <- prcomp(winter)
summary(winterpca)
winterpca$loadings[, 1:2]

fviz_eig(winterpca, addlabels = TRUE)

fviz_pca_var(winterpca, col.var = "black")

fviz_cos2(winterpca, choice = "var", axes = 1:2)
```

```{r sixweek}
omittedsixweek <- na.omit(six_week_data)
sixweek <- select_if(omittedsixweek, is.numeric)
sixweek <- subset(sixweek, select = -Gender)
sixweek <- scale(sixweek)
sixweekpca <- prcomp(sixweek)
summary(sixweekpca)
sixweekpca$loadings[, 1:2]

fviz_eig(sixweekpca, addlabels = TRUE)

fviz_pca_var(sixweekpca, col.var = "black")

fviz_cos2(sixweekpca, choice = "var", axes = 1:2)
```

```{r summervisual}
summercomps <- as.data.frame(summerpca$x)
newsummer <- cbind(omittedsummer,summercomps[,c(1,2)])
head(omittedsummer)
ggplot(newsummer, aes(x=PC1, y=PC2, col = Ethnicity, fill = Ethnicity)) +
  stat_ellipse(geom = "polygon", col= "black", alpha =0.5)+
  geom_point(shape=21, col="black") +
  labs(title = "Summer PCA")

fviz_contrib(summerpca, choice = "var", axes = 1, top = 20)
fviz_contrib(summerpca, choice = "var", axes = 2, top = 20)
```

```{r wintervisual}
wintercomps <- as.data.frame(winterpca$x)
newwinter <- cbind(omittedwinter,wintercomps[,c(1,2)])
ggplot(newwinter, aes(x=PC1, y=PC2, col = Ethnicity, fill = Ethnicity)) +
  stat_ellipse(geom = "polygon", col= "black", alpha =0.5)+
  geom_point(shape=21, col="black") + 
  labs(title = "Winter PCA")

fviz_contrib(winterpca, choice = "var", axes = 1, top = 20)
fviz_contrib(winterpca, choice = "var", axes = 2, top = 20)
```

```{r sixweekvisual}
sixweekcomps <- as.data.frame(sixweekpca$x)
newsixweek <- cbind(omittedsixweek,sixweekcomps[,c(1,2)])
ggplot(newsixweek, aes(x=PC1, y=PC2, col = Ethnicity, fill = Ethnicity)) +
  stat_ellipse(geom = "polygon", col= "black", alpha =0.5)+
  geom_point(shape=21, col="black") + 
  labs(title = "Six Week PCA")

fviz_contrib(sixweekpca, choice = "var", axes = 1, top = 20)
fviz_contrib(sixweekpca, choice = "var", axes = 2, top = 20)
```

```{r vitdgradient}
ggplot(newsummer, aes(x=PC1, y=PC2, col = VitDResult, fill = VitDResult)) +
  geom_point(aes(shape = Ethnicity), size = 3, stroke = 1, col = "black") + 
  scale_fill_gradient(low="blue", high="red") + 
  labs(title = "Summer PCA") +
  theme_minimal() +
  scale_shape_manual(values = c(21, 22)) 

ggplot(newwinter, aes(x=PC1, y=PC2, col = VitDResult, fill = VitDResult)) +
  geom_point(aes(shape = Ethnicity), size = 3, stroke = 1, col = "black") + 
  scale_fill_gradient(low="blue", high="red") + 
  labs(title = "Summer PCA") +
  theme_minimal() +
  scale_shape_manual(values = c(21, 22)) 

ggplot(newsixweek, aes(x=PC1, y=PC2, col = VitDResult, fill = VitDResult)) +
  geom_point(aes(shape = Ethnicity), size = 3, stroke = 1, col = "black") + 
  scale_fill_gradient(low="blue", high="red") + 
  labs(title = "Summer PCA") +
  theme_minimal() +
  scale_shape_manual(values = c(21, 22)) 
```


# part 2: long data PCA

```{r long_pca}
omittedlong <- na.omit(long_data)
long <- select_if(omittedlong, is.numeric)
long <- subset(long, select = -Gender)
long <- scale(long)
longpca <- prcomp(long)
summary(longpca)
longpca$loadings[, 1:2]

fviz_eig(longpca, addlabels = TRUE)

fviz_pca_var(longpca, col.var = "black")

fviz_cos2(longpca, choice = "var", axes = 1:2)
```

```{r summervis}
longcomps <- as.data.frame(longpca$x)
newlong <- cbind(omittedlong,longcomps[,c(1,2)])
head(omittedlong)
ggplot(newlong, aes(x=PC1, y=PC2, col = Ethnicity, fill = Ethnicity)) +
  stat_ellipse(geom = "polygon", col= "black", alpha =0.5)+
  geom_point(shape=21, col="black") +
  labs(title = "Summer PCA")

fviz_contrib(longpca, choice = "var", axes = 1, top = 20)
fviz_contrib(longpca, choice = "var", axes = 2, top = 20)
```

```{r longgradient}
pal <- wes_palette("Zissou1", 10, type = "continuous")
ggplot(newlong, aes(x=PC1, y=PC2, col = VitDResult, fill = VitDResult)) +
  geom_point(aes(shape = Ethnicity), size = 3, stroke = 1, col = "black") + 
  scale_fill_gradientn(colors = pal) +
  labs(title = "Long PCA") +
  theme_minimal() +
  scale_shape_manual(values = c(21, 22)) 
```

```{r widepca}
omittedwide <- na.omit(joined_data)
wide <- select_if(omittedwide, is.numeric)
wide_standard <- scale(wide)
widepca <- prcomp(wide_standard, center = T, scale. = T)
summary(widepca)
widepca$rotation
widepca_scores <- widepca$x

widecomps <- as.data.frame(widepca$x)
newwide <- cbind(omittedwide,widecomps[,c(1,2)])
head(omittedwide)
ggplot(newwide, aes(x=PC1, y=PC2, col = 'Ethnicity-summer', fill = 'Ethnicity-summer')) +
  stat_ellipse(geom = "polygon", col= "black", alpha =0.5)+
  geom_point(shape=21, col="black") +
  labs(title = "Wide PCA")
```








